---
title: "Introduction to model fitting in Julia"
format: html
engine: julia
---

[Lecture slides](slides/intro_slides.pdf)

# Introduction

# Objectives

::: {.callout-note collapse="true"}

# Setup

```{julia}
#| echo: false
using Pkg
Pkg.activate("..")
```

## Source file

The source file of this session is located at `sessions/introduction.qmd`.

## Load libraries

```{julia}
using Random ## for random numbers
using Turing ## for probabilistic modelling
using Distributions ## for probability distributions
using DifferentialEquations ## for differential equations
using DataFrames ## for data frames
using Plots ## for plots
using StatsPlots ## for statistical plots
```

## Initialisation

We set a random seed for reproducibility. 
Setting this ensures that you should get exactly the same results on your computer as we do.

```{julia}
Random.seed!(1234)
```

:::

The aim of this first session is to set you up with a framework for model fitting using simulation-based intuition. We will use **Turing.jl** to define models and **Distributions.jl** for elegant statistical operations, but we'll start by building understanding through hands-on simulation rather than jumping straight into automated inference.

In this session, you will

1. define a Bayesian model using **Turing.jl** syntax
2. understand **priors**, **likelihood**, and **posterior** through simulation
3. manually explore the posterior by trying different parameter values
4. understand why we need algorithmic approaches (setting up the next session)

The key insight: Bayesian inference is fundamentally about simulation and comparison. 
We will do this manually first, then automate it later.

# A deterministic SIR model in Julia

We'll work with a simple deterministic SIR model. The model has two parameters: the basic reproductive number $R_0$ and the infectious period $D_\mathrm{inf}$. The model equations are:

$$
\begin{cases}
\begin{aligned}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - \beta S \frac{I}{N}\\
\frac{\mathrm{d}I}{\mathrm{d}t} &= \beta S \frac{I}{N} - \nu I\\
\frac{\mathrm{d}R}{\mathrm{d}t} &= \nu I\\
\end{aligned}
\end{cases}
$$

where $\beta=R_0/D_\mathrm{inf}$ and $\nu = 1/D_\mathrm{inf}$.

To use this model in Julia we first define a function that represents the ODEs:

```{julia}
# Define ODE function
function sir_ode!(du, u, p, t)
    S, I, R = u
    R_0, D_inf = p
    N = S + I + R
        
    β = R_0 / D_inf
    ν = 1 / D_inf

    du[1] = -β * S * I / N  # dS/dt
    du[2] = β * S * I / N - ν * I  # dI/dt
    du[3] = ν * I  # dR/dt
end
```

::: {.callout-tip collapse = "true"}
## Why is there an exclamation mark in `sir_ode!()`?

In Julia, the exclamation mark indicates that the function can modify its arguments.
In this case, the function modifies `du` in the last 3 lines to indicate the current change in the system as given by the ODEs.
:::

In order to simulate this model, i.e. numerically integrate the differential equations, we define it as an `ODEproblem` (which tells Julia that the function we defined above represents the dynamics of the system) which we then solve:

```{julia}
#| output: false

## define initial state
init_state = [999, 1, 0]
## define start and end times
t0, tend = (0, 30)
## define parameters: R_0 = 2.5, D_inf = 2
p = [2.5, 2]
# Define ODE problem
prob = ODEProblem(sir_ode!, init_state, (t0, tend), p)
 # Solve ODE
sol = solve(prob, saveat = 1)
```

::: {.callout-tip collapse = "true"}
## More details on these commands
Let's take a closer look at what each of these commands do:
```{{julia}}
init_state = [999, 1, 0]
```
This creates an 1-dimensional [Array](https://docs.julialang.org/en/v1/manual/arrays/) (i.e., a vector) with three elements that we will later use to represent the initial state of our forward simulation (999 susceptibles, 1 infective and 0 recovered).
The output tells us that the elements of the vector are of type `Int64`, i.e. integer-valued numbers.

```{{julia}}
t0, tend = (t0, tend)
```
This assigns values to the variables `t0` (0) and `tend` (30).
It is just a shorter version of writing
```{{julia}}
t0 = 0
tend = 40
```

```{{julia}}
prob = ODEProblem(sir_ode!, init_state, (t0, tend), p)
```

This defines our ODE system. 
The first argument to `ODEproblem()` is a function that contains the equations of the ODE, in our case the `sir_ode!()` function we defined earlier.
The second argument is the initial state.
The third argument defines that start and end time of the simulation, here given as a _tuple_ indicated by round parentheses, `(t0, tend)`.
A tuple is similar to an array (defined with brackets as we did above in defining `init_state`) but more flexible in allowing different data types, as in e.g. lists in R.
Here we could also have used a vector `[t0, tend]`.
Either will be correctly interpreted by `ODEProblem()`.
The fourth argument gives the parameters of the model `p` as a vector.

```{{julia}}
sol = solve(prob, saveat = 1)
```

This solves the ODE problem given by `prob`.
The `saveat` argument tells the solver to save state information at time steps of size 1.
If this is not given, the time steps at which information is saved will be chosen by the ODE solver - this is often fine but as we will later want to compare this to observations at integer time steps we here save information at these distinct steps.
:::

The `sol` object now contains information on the system dynamics (contained in `sol.u`) at different times (contained in `sol.t`).

We can plot this using

```{julia}
plot(sol, labels=["S" "I" "R"])
```

# A full Bayesian model in Turing.jl

We can use the _simulation model_ we defined above to define a full Bayesian model including prior and likelihood distributions:

```{julia}
#| output: false
#
@model function sir_model(df, init_state)
    # Priors
    R_0 ~ Uniform(1, 100)
    D_inf ~ Uniform(0.1, 30)

    times = df.time
    # Process model
    prob = ODEProblem(
        sir_ode!, init_state, (times[1], times[end]), [R_0, D_inf]
    )
    ## save daily 
    sol = solve(prob, saveat = 1)

    data = df.obs
    # Likelihood
    for i in 1:length(data)
        ## fit to incidence data: newly infected given by change in S
        incidence = sol.u[times[i] - 1][1] - sol.u[times[i]][1]
        data[i] ~ Poisson(incidence)
    end
end
```

This is the complete Bayesian model.
In the next few sections we will go through it bit by bit to develop an understanding of its components.

# Part 1: Understanding priors

## What is a prior?

A **prior** represents what we believe about parameters **before** seeing any data. It encodes our background knowledge, expertise, or assumptions about plausible parameter values.

In our SIR model:
- $R_0 \sim \text{Uniform}(1, 100)$: We believe $R_0$ could be anywhere from 1 to 100, with all values equally likely
- $D_\mathrm{inf} \sim \text{Uniform}(0.1, 30)$: We believe infectious periods from 0.1 to 30 days are equally plausible

Let's explore what these priors mean through simulation:

```{julia}
# Define our priors as Distribution objects
R_0_prior = Uniform(1.0, 100.0)
D_inf_prior = Uniform(0.1, 30.0)

# Sample from priors many times
n_samples = 1000
R_0_samples = rand(R_0_prior, n_samples)
D_inf_samples = rand(D_inf_prior, n_samples)

println(
    "R_0 prior samples - mean: ", round(mean(R_0_samples), digits = 2), 
    ", range: ", round(minimum(R_0_samples), digits = 2), " to ",
    round(maximum(R_0_samples), digits = 2)
)
println(
    "D_inf prior samples - mean: ", round(mean(D_inf_samples), digits = 2), 
    ", range: ", round(minimum(D_inf_samples), digits = 2), " to ",
    round(maximum(D_inf_samples), digits = 2)
)
```

```{julia}
# Visualize our priors
p1 = histogram(
    R_0_samples, title = "R_0 Prior", xlabel = "R_0", ylabel = "Frequency",
    alpha = 0.7, label = "Uniform(1, 100)"
)
p2 = histogram(
    D_inf_samples, title = "D_inf Prior", xlabel = "D_inf (days)",
    ylabel = "Frequency", alpha = 0.7, label = "Uniform(0.1, 30)"
)
plot(p1, p2, layout = (1,2))
```

::: {.callout-note collapse = "true"}
### Coming from R?
You might be familiar with different functions for calculating probability densities (e.g., `dnorm()`) or generate random numbers (e.g. `rnorm()` for generating random numbers for given probability distributions.
In Julia we can use the interface provided by [Distributions.jl](https://juliastats.org/Distributions.jl/stable/) to define the probability distribution and then calculate probability densities or generate random variates:
```julia
## define normal distribution with mean 3 and standard deviation 0.1
dist = Normal(3, 0.1)
## generate 10 random numbers from this distribution
Rand(dist, 10)
## calculate the (logged) probability distribution at x = 2
logdpf(dist, 2)
```

The equivalent R code would be
```{r eval = FALSE}
rnorm(10, 3, 0.1)
dnorm(2, 3, 0.1)
```
:::

## Prior predictive simulation

An important question is: **What trajectories do our priors represent?** 
Let's find out by simulating epidemics using random draws from our priors:

```{julia}
# Prepare plot where we'll add the individual trajectories
p = plot(
    title = "Prior Predictive Trajectories",
    xlabel = "Time", ylabel = "Infected"
)

ntraj = 20

for i in 1:ntraj
    R_0 = R_0_samples[i]
    D_inf = D_inf_samples[i]

    prob = ODEProblem(
        sir_ode!, init_state, (t0, tend), [R_0, D_inf]
    )
    sol = solve(prob)
    plot!(p, sol, alpha = 0.6, label = "", color = :grey, idxs = (0, 2))
end

plot!(p)
```

::: {.callout-note collapse = "true"}
### How did we tell Julia to plot just the number of infected?
The `plot()` commands now contain an argument `idxs`, which allows us to choose what to plot.
This uses the [plot recipe for differential equations](https://docs.sciml.ai/DiffEqDocs/stable/basics/plot).
Index 0 corresponds to the time variable, and any indices >0 to the variables of the differential equations.
In our case, we specify `0, 2`, that is we plot time (on the x axis) against the second variable, I (on the y axis).
:::

**Take 2 minutes**: Look at these trajectories.
Do they seem reasonable? Are there any that look unrealistic? 

Some questions to consider:

- Do any grow too fast or too slow?
- Are the peak sizes reasonable for a population of 1000?
- Do any last too long or too short?

This is **prior predictive checking** - a crucial step to validate that our priors make sense.

## Choosing different priors

What if we had different beliefs?
Let's try more informative (i.e., narrower) priors:

```{julia}
# More informative priors based on epidemiological knowledge
R_0_informed = Normal(2.5, 0.8)  # Most respiratory diseases have R_0 around 2-3
D_inf_informed = Gamma(2, 1)     # Infectious period likely 1-4 days for common infections

# Sample and visualize
R_0_informed_samples = rand(R_0_informed, 1000)
D_inf_informed_samples = rand(D_inf_informed, 1000)

p1 = histogram(
    R_0_samples, title = "Flat Prior", xlabel = "R_0", alpha = 0.7,
    label = "Uniform(1, 100)"
)
histogram!(p1, R_0_informed_samples, alpha = 0.7, label = "Normal(2.5, 0.8)")

p2 = histogram(
    D_inf_samples, title = "Flat Prior", xlabel = "D_inf", alpha = 0.7,
    label = "Uniform(0.1, 30)"
)
histogram!(p2, D_inf_informed_samples, alpha = 0.7, label = "Gamma(2, 1)")

plot(p1, p2, layout = (1,2))
```

```{julia}
# Compare epidemic predictions
p = plot(
    title = "Prior Predictive: Flat vs Informed",
    xlabel = "Time", ylabel = "Infected"
)

# Flat priors (gray)
for i in 1:10
    R_0 = R_0_samples[i]
    D_inf = D_inf_samples[i]

    prob = ODEProblem(
        sir_ode!, init_state, (t0, tend), [R_0, D_inf]
    )
    sol = solve(prob)
    plot!(p, sol, alpha = 0.6, label = "", color = :gray, idxs = (0, 2))
end

# Informed priors (blue)
for i in 1:10
    R_0 = R_0_informed_samples[i]
    D_inf = D_inf_informed_samples[i]

    prob = ODEProblem(
        sir_ode!, init_state, (t0, tend), [R_0, D_inf]
    )
    sol = solve(prob)
    plot!(p, sol, alpha = 0.6, label = "", color = :blue, idxs = (0, 2))
end

plot!(p)
```

**Key insight**: Priors determine what kinds of epidemics our model considers plausible before seeing data.

# Part 2: Understanding the likelihood

## What is a likelihood?

The **likelihood** quantifies how probable our observed data is, given specific parameter values.
It asks: "If the parameters were X, how likely would we be to see the data we actually observed?"

The predictive prior trajectories we looked at earlier were purely a function of the model and its specification.
The likelihood, on the other hand, is about comparing the model to data.
Let's start by loading some observed epidemic data (we simulated these from the model but in principle this could be any data) representing the number of infectious individuals at different time points:

```{julia}
# Define data frame with epidemic  data
epi1 = DataFrame(
    time = 1:1:30, ## 30 data points
    obs = [
        1, 3, 4, 11, 19, 38, 67, 112, 183, 210, 237, 182, 178, 154, 95, 68, 51,
        30, 32, 18, 19, 9, 2, 2, 2, 0, 0, 1, 0, 0
    ]
)

println("Data overview:")
first(epi1, 8)
```

We first visualise this data set.

```{julia}
# Visualise the observed data
scatter(
    epi1.time, epi1.obs, xlabel = "Time", ylabel = "Observed Cases", 
    title = "Epidemic Data", label = "Observed", markersize = 4, color = :red
)
```

This data was generated from an SIR model with $R_0 ≈ 2.5$ and $D_\mathrm{inf} = 2$ days.
Let's see if we could have figured this out by exploring the likelihood.

## Exploring likelihood through simulation

The likelihood asks: "How probable is our observed data, given specific parameter values?"
Let's explore this by trying different parameter values:

```{julia}
# Try a few different parameter combinations
test_params = [
    (R_0 = 1.1, D_inf = 2.0),
    (R_0 = 1.5, D_inf = 2.0),
    (R_0 = 2.5, D_inf = 2.0),
    (R_0 = 4.0, D_inf = 2.0),
    (R_0 = 2.5, D_inf = 1.0),
    (R_0 = 2.5, D_inf = 4.0)
]

# Generate plot
p = plot(title = "Model Fit Comparison", xlabel = "Time", ylabel = "Cases")

# For each parameter set, calculate log-likelihood and plot
for theta in test_params

    R_0 = theta[:R_0]
    D_inf = theta[:D_inf]

    # Simulate from the model
    prob = ODEProblem(
        sir_ode!, init_state, (t0, tend), [R_0, D_inf]
    )
    sol = solve(prob, saveat = epi1.time)

    ## I compartment is second compartment
    I = sol[2, :]
    ## assuming Poisson log-likelihood at each time point
    loglik = [ Poisson(It) for It in I ]
    ## calculate loglikelihood as sum log-likelihoods at each time point
    total_loglik = sum(logpdf.(loglik, epi1.obs))
    println(
        "R_0 = $(R_0), D_inf = $(D_inf) → ",
        "log-liklihood  =  $(round(total_loglik, digits = 1))"
    )
    plot!(
        p, sol, idxs = (0, 2),
        label = "R_0 = $(theta.R_0), D_inf = $(theta.D_inf)", linewidth = 2
    )
end

scatter!(
    p, epi1.time, epi1.obs, label = "Observed", markersize = 4, color = :black,
    markershape = :circle
)
plot!(p)
```

**Take 5 minutes**: Which parameter combination gives the highest likelihood (least negative log-likelihood)? Which one visually fits the data best?


## Understanding the likelihood function

Our likelihood assumes each observation follows a Poisson distribution around the true prevalence. Let's understand this:

```{julia}
# Pick one time point and explore the likelihood in detail
time_point = 10
theta_example = (R_0 = 2.5, D_inf = 2.0)
prob = ODEProblem(
    sir_ode!, init_state, (t0, tend), [theta_example.R_0, theta_example.D_inf]
)
sol = solve(prob, saveat = epi1.time)

expected_cases = sol[2,:][time_point]
observed_cases = epi1.obs[time_point]

println("At time $(epi1.time[time_point]):")
println("  Expected cases: $(round(expected_cases, digits=1))")
println("  Observed cases: $observed_cases")

# Show the Poisson distribution
obs_dist = Poisson(expected_cases)
possible_observations = 0:500
probabilities = pdf.(obs_dist, possible_observations)

p = plot(
    possible_observations, probabilities,
    title = "Likelihood for one data point", 
    xlabel = "Possible observations", ylabel = "Probability", linewidth = 2
)
vline!(
    p, [observed_cases], color = :red, linewidth = 3,
    label = "Actual observation"
)
scatter!(
    p, [observed_cases], [pdf(obs_dist, observed_cases)], color = :red,
    markersize = 8, label = "Likelihood value"
)
```

**Key insight**: The likelihood is highest when our model prediction matches the observation closely.

::: {.callout-note title="Coming from R?"}
In R, you would calculate Poisson log-likelihood as:
```{r eval = FALSE}
log_lik <- dpois(observed, lambda = expected, log=TRUE)
```

In Julia instead we have used with Distribution objects:
```julia
obs_dist = Poisson(expected)
logpdf(obs_dist, observed)
```

This same `obs_dist` object can be used for random sampling (`rand(obs_dist)`) and other operations.
:::

# Part 3: Understanding the posterior

## What is a posterior?

The **posterior** combines our prior beliefs with the data evidence using **Bayes' theorem**:

$$p(\theta|\text{data}) \propto p(\text{data}|\theta) \times p(\theta)$$

- $p(\theta|\text{data})$: **Posterior** - what we believe about parameters *after* seeing data
- $p(\text{data}|\theta)$: **Likelihood** - how well parameters explain the data
- $p(\theta)$: **Prior** - what we believed *before* seeing data

In log space: $\log p(\theta|\text{data}) = \log p(\text{data}|\theta) + \log p(\theta) + \text{constant}$

Let's implement this step by step:

```{julia}
# Test posterior calculation using Turing.jl conditioning
for theta in test_params
    # Calculate components separately for educational purposes
    log_prior_val = logprior(model, theta)
    log_lik = loglikelihood(model, theta)
    log_post = logjoint(model, theta)  # This should equal log_prior_val + log_lik
    
    println("R_0=$(theta.R_0), D_inf=$(theta.D_inf)")
    println("  Prior: $(round(log_prior_val, digits=1)), Likelihood: $(round(log_lik, digits=1)), Posterior: $(round(log_post, digits=1))")
    println()
end
```

## How priors influence the posterior

Let's see how the choice of prior affects our conclusions:

```{julia}
# Compare flat vs informative priors for the same parameter
theta_test = (R_0 = 2.5, D_inf = 2.0)

# Flat prior (what we used above)
flat_R0_prior = logpdf(R_0_prior, 2.5)
flat_lik = loglikelihood(model, theta_test)
flat_post = logjoint(model, theta_test)

# Informative prior that favors R_0 ≈ 2.5
informed_R0_prior = logpdf(Normal(2.5, 0.3), 2.5)  # Strong belief R_0 is near 2.5
informed_post = flat_lik + informed_R0_prior + logpdf(D_inf_prior, 2.0)

println("At R_0=2.5, D_inf=2.0:")
println("Flat prior: $(round(flat_R0_prior, digits=2))")
println("Informed prior: $(round(informed_R0_prior, digits=2))")
println("Likelihood: $(round(flat_lik, digits=1))")
println()
println("Flat posterior: $(round(flat_post, digits=1))")
println("Informed posterior: $(round(informed_post, digits=1))")
println("Difference: $(round(informed_post - flat_post, digits=1))")
```

**Key insight**: When we have strong prior beliefs that align with the data, the posterior gets a boost. When priors conflict with data, they pull the posterior away from the likelihood.

# Manual posterior exploration

Let's systematically explore the posterior by trying many parameter combinations:

```{julia}
# Grid search over parameter space using Turing.jl conditioning
R_0_values = 1.5:0.2:4.0
D_inf_values = 1.0:0.2:3.0

best_log_posterior = -Inf
best_params = nothing
results = []

println("Exploring posterior... (this may take a moment)")

for R_0 in R_0_values
    for D_inf in D_inf_values
        theta = (R_0=R_0, D_inf=D_inf)
        # Calculate log-posterior
        log_post = logjoint(model, theta)
        push!(results, (R_0=R_0, D_inf=D_inf, log_posterior=log_post))
        
        if log_post > best_log_posterior
            best_log_posterior = log_post
            best_params = theta
        end
    end
end

println("Best parameters found: R_0 = $(best_params.R_0), D_inf = $(best_params.D_inf)")
println("Best log-posterior: $(round(best_log_posterior, digits=1))")
```

```{julia}
# Visualize the posterior surface
R_0_grid = [r.R_0 for r in results]
D_inf_grid = [r.D_inf for r in results]
log_post_grid = [r.log_posterior for r in results]

# Reshape for heatmap
R_0_unique = unique(R_0_grid)
D_inf_unique = unique(D_inf_grid)
post_matrix = reshape(log_post_grid, length(R_0_unique), length(D_inf_unique))

heatmap(D_inf_unique, R_0_unique, post_matrix, 
        xlabel="D_inf (days)", ylabel="R_0", 
        title="Log-Posterior Surface", color=:viridis)
```

## The best fit

Now let's see how well our best parameters fit:

```{julia}
# Plot the best fit
best_trajectory = sir_simulate_deterministic(best_params, init_state, epi1.time)

plot(sol, idxs = (0, 2),
     label="Best fit (R_0=$(best_params.R_0), D_inf=$(best_params.D_inf))", 
     linewidth=3, title="Best Manual Fit")
scatter!(epi1.time, epi1.obs, label="Observed", markersize=4, color=:red)
xlabel!("Time")
ylabel!("Cases")
```

**Great!** But notice some problems with our manual approach:

1. **Grid search is expensive**: We only tried `$(length(R_0_values) × length(D_inf_values))` combinations
2. **No uncertainty quantification**: We found the "best" parameters, but how certain are we?
3. **Doesn't scale**: What if we had 5 parameters? 10 parameters?
4. **Coarse resolution**: We might be missing the true optimum between grid points

# Understanding uncertainty

Let's get a sense of parameter uncertainty by looking at all "reasonably good" parameter combinations:

```{julia}
# Define "reasonably good" as within 10 log-likelihood units of the best
threshold = best_log_posterior - 10
good_results = filter(r -> r.log_posterior > threshold, results)

println("Found $(length(good_results)) 'good' parameter combinations out of $(length(results)) total")

# Plot posterior uncertainty
p = plot(title="Parameter Uncertainty", xlabel="Time", ylabel="Cases")

for result in good_results[1:min(50, end)]  # Plot up to 50 trajectories
    theta = (R_0=result.R_0, D_inf=result.D_inf)
    trajectory = sir_simulate_deterministic(theta, init_state, epi1.time)
    plot!(p, trajectory.time, trajectory.I, alpha=0.3, color=:blue, label="")
end

# Highlight the best fit
plot!(p, best_trajectory.time, best_trajectory.I, 
      linewidth=3, color=:red, label="Best fit")
scatter!(p, epi1.time, epi1.obs, label="Observed", markersize=4, color=:black)
plot!(p)
```

```{julia}
# Look at the distribution of good parameters
good_R0 = [r.R_0 for r in good_results]
good_D_inf = [r.D_inf for r in good_results]

p1 = histogram(good_R0, title="Uncertain R_0", xlabel="R_0", alpha=0.7)
vline!(p1, [best_params.R_0], linewidth=3, color=:red, label="Best")

p2 = histogram(good_D_inf, title="Uncertain D_inf", xlabel="D_inf", alpha=0.7)
vline!(p2, [best_params.D_inf], linewidth=3, color=:red, label="Best")

plot(p1, p2, layout=(1,2), size=(800,300))
```

**This is getting closer to real Bayesian inference!** We're seeing the uncertainty in our parameter estimates.

# Model checking through simulation

One way to check our model is to generate fake data using our best parameters and see if it looks like the real data:

```{julia}
# Generate several datasets using our best parameters
p = plot(title="Posterior Predictive Check", xlabel="Time", ylabel="Cases")

for i in 1:20
    # Simulate the trajectory
    trajectory = sir_simulate_deterministic(best_params, init_state, epi1.time)
    
    # Generate noisy observations following our likelihood model
    fake_obs = [rand(Poisson(max(trajectory.I[j], 1e-6))) for j in 1:nrow(epi1)]
    
    plot!(p, epi1.time, fake_obs, alpha=0.4, color=:lightblue, label="")
end

# Add the real data
scatter!(p, epi1.time, epi1.obs, label="Real data", markersize=4, color=:red)
plot!(p)
```

**Take 5 minutes**: Does the real data look consistent with our simulated datasets? This is **posterior predictive checking** - a key model validation technique.

# Why we need algorithmic approaches

Through simulation, we've built intuition about:
- **Priors**: Our beliefs before seeing data
- **Likelihood**: How well parameters explain the data  
- **Posterior**: Combining beliefs with evidence
- **Uncertainty**: Multiple parameter values are consistent with data

But our manual approach has serious limitations:

```{julia}
# The curse of dimensionality
n_params = [2, 3, 4, 5, 6, 10]
n_values_per_param = 15  # Reasonable resolution

println("Grid search scaling problems:")
for n in n_params
    total_evals = n_values_per_param^n
    if total_evals < 1e6
        println("$(n) parameters: $(total_evals) evaluations")
    else
        println("$(n) parameters: $(round(total_evals/1e6, digits=1)) million evaluations")
    end
end
```

**Problems with manual exploration:**

1. **Computational cost**: Grid search doesn't scale
2. **Poor sampling**: We might miss important regions
3. **No convergence guarantees**: How do we know when we've explored enough?
4. **Inefficient**: We waste time in low-probability regions

**Next session**: We'll learn how Markov Chain Monte Carlo (MCMC) algorithms solve these problems by automatically exploring the posterior distribution. We'll start with simple algorithms you can code yourself, then see how Turing.jl automates everything with state-of-the-art methods.

# Summary

In this session, you learned:

1. **Priors**: Express beliefs before seeing data, explored through simulation
2. **Likelihood**: Quantifies how well parameters explain observed data
3. **Posterior**: Combines priors and likelihood using Bayes' theorem
4. **Manual exploration**: Grid search to find best parameters and understand uncertainty
5. **Model checking**: Using simulation to validate model assumptions
6. **Limitations**: Why manual approaches don't scale

**Key insights**:
- **Bayesian inference is about updating beliefs** with data
- **Simulation builds intuition** for abstract concepts
- **Uncertainty quantification** is as important as point estimates
- **Manual approaches teach concepts** but don't scale

**Coming up**: Automated posterior exploration with MCMC algorithms!
