---
title: "Example: fitting the deterministic SEITL model"
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(cache = TRUE, cache.path = "cache/example_mcmc_SEITL/", fig.retina = 2, fig.path = "figure/example_mcmc_SEITL/", fig.cap = "", fig.align = "center", message = FALSE, tidy = TRUE, warning = FALSE, fig.width = 7, fig.height = 7)
```

```{r fitR, cache = FALSE, echo = FALSE, results = "hide", warning = FALSE}
library(fitR)
set.seed(1234)
```

Here is an example of possible answers to the practical on fitting the deterministic SEITL model to the Tristan da Cunha outbreak. 

Each section below correspond to a section of the practical. Thus, you can have a look at our example for one section and then go back to the practical to answer the following sections. 

Although our example refers to the SEITL model, the same commands work for the SEIT4L model (i.e. `data(seit4lDeter)` instead of `data(seitlDeter)`).

# Setting the MCMC

```{r set-mcmc, eval = FALSE}
# the fitmodel
data(seitlDeter)

# wrapper for posterior
my_posteriorTdC <- function(theta) {
  my_fitmodel <- seitlDeter
  my_initState <- c("S" = 279, "E" = 0, "I" = 2, "T" = 3, "L" = 0, "Inc" = 0)
  # note that for the SEIT4L model there are 4 state variables for the T compartment
  # my_initState <- c("S" = 279, "E" = 0, "I" = 2, "T1" = 3, "T2" = 0, "T3" = 0, "T4" = 0, "L" = 0, "Inc" = 0)

  return(dLogPosterior(fitmodel = my_fitmodel, theta = theta, initState = my_initState, data = FluTdC1971, margLogLike = dTrajObs, log = TRUE))
}

# theta to initialise the MCMC
initTheta <- c("R0" = 2, "D_lat" = 2, "D_inf" = 2, "alpha" = 0.8, "D_imm" = 16, "rho" = 0.85)

# diagonal elements of the covariance matrix for the Gaussian proposal
proposalSd <- c("R0" = 1, "D_lat" = 0.5, "D_inf" = 0.5, "alpha" = 0.1, "D_imm" = 2, "rho" = 0.1)

# lower and upper limits of each parameter
lower <- c("R0" = 0, "D_lat" = 0, "D_inf" = 0, "alpha" = 0, "D_imm" = 0, "rho" = 0)
upper <- c("R0" = Inf, "D_lat" = Inf, "D_inf" = Inf, "alpha" = 1, "D_imm" = Inf, "rho" = 1)

# number of iterations for the MCMC
nIterations <- 5000

# additional parameters for the adaptive MCMC, see ?mcmcMh for more details
adaptSizeStart <- 100
adaptSizeCooling <- 0.999
adaptShapeStart <- 200
```

You can now go back to the [practical](mcmc_and_model_comparison.html#run-a-mcmc) and try to run MCMC with those settings.

# Run MCMC

If you didn't manage to run MCMC, or it took too long to obtain a few thousand iterations, you can load our short run as follows:

```{r load-trace, collapse = TRUE}
data(mcmc_TdC_deter_shortRun)
# this should load 2 objects in your environment: mcmc_SEITL and mcmc_SEIT4L. Each one is a list of 3 elements returned by mcmcMh
names(mcmc_SEITL)
# the trace contains 9 variables for 5000 iterations
dim(mcmc_SEITL$trace)
# let's have a look at it
head(mcmc_SEITL$trace)
```
You can now go back to the [practical](mcmc_and_model_comparison.html#short-run-analysis) and analyse this trace.

# Short run analysis

Here is an example of analysis for our preliminary run:

```{r short-run-analysis-trace, collapse = TRUE}
# convert to a mcmc object for coda
library("coda")
trace <- mcmc(mcmc_SEITL$trace)

# compute the acceptance rate
1 - rejectionRate(trace)
# between 0.1 and 0.6: looks good!

# let's have a look at the traces
library("lattice") ## for the 'xyplot' command
xyplot(x = trace)
```

Although the chain was started at a `initTheta` with a low posterior density, it quickly finds the region of the parameter space with high posterior density. Note also the constant trace of the log-prior since we have assumed a uniform prior.

Overall, it looks like the chain reached its target distribution after 1000 steps.

```{r short-run-analysis-burn-test, collapse = TRUE}
# Let's find a suitable burning:
plotEssBurn(trace)
```

As anticipated from the trace, burning the first 1000 iterations maximizes the effective sample size (ESS).

```{r short-run-analysis-burn, collapse = TRUE}
# Let's create a new trace without the burning
traceBurn <- burnAndThin(trace, burn = 1000)
xyplot(x = traceBurn)
# Let's check the ESS
effectiveSize(traceBurn)
```
Although we have 4000 samples remaining after burning, the ESS is much smaller. This is due to autocorrelation of the chain.

```{r short-run-analysis-acf}
# autocorrelation plot
acfplot(x = traceBurn, lag.max = 60)
```
The autocorrelation between samples drops substantially for a lag of 20 iterations. We can thin the trace to reduce the autocorrelation.

```{r short-run-analysis-thin, collapse = TRUE}
# Let's create a thinned trace
traceBurnThin <- burnAndThin(traceBurn, thin = 20)
xyplot(x = traceBurnThin)
# Let's check the ESS
effectiveSize(traceBurnThin)
```
Although the thinned trace has 20 times less fewer than the unthinned trace, it has a similar ESS. This is because the autocorrelation has been reduced.

```{r short-run-analysis-acf-thin}
# new autocorrelation plot
acfplot(x = traceBurnThin, lag.max = 60)
```

Let's compare the posterior estimates of the thinned and unthinned traces.

```{r short-run-analysis-compare-summary, collapse = TRUE}
# The unthinned trace
summary(traceBurn)

# The thinned trace
summary(traceBurnThin)
```

They are very similar. So why thin? Because autocorrelation produces clumpy samples that are unrepresentative, in the short run, of the true underlying posterior distribution. We can check this by comparing the thinned and unthinned distributions using the function `plotPosteriorDensity` of the `fitR` package:

```{r short-run-analysis-compare-density, fig.width=10}
plotPosteriorDensity(list(unthinned = traceBurn, thinned = traceBurnThin))
```
The thinned trace shows a smoother distribution despite having less samples than the unthinned one. This because the local "bumps" of the unthinned distribution are caused by autocorrelated samples.

You can now go back to the [practical](mcmc_and_model_comparison.html#long-run-analysis) and perform a similar analysis for a long-run MCMC.

# Long run analysis

Here is an example of an analysis for our long run ($10^5$ iterations)

```{r long-run-combine-traces, collapse = TRUE}
# load mcmc output
data(mcmc_TdC_deter_longRun)

# create mcmc objects for both traces
library("coda")
trace1 <- mcmc(mcmc_SEITL_theta1$trace)
trace2 <- mcmc(mcmc_SEITL_theta2$trace)

# combine traces as mcmc.list object
trace <- mcmc.list(list(trace1, trace2))

# let's have a look
head(trace, 3)

# acceptance rate
1 - rejectionRate(trace)
# close to the optimal value of 0.234

# ESS
effectiveSize(trace)

# plot the traces
library("lattice") ## for the 'xyplot' command
xyplot(trace)
```
Note that the acceptance rate and the ESS are computed for the combined chain whereas the traces are plotted for each chain.
Also, given the very high ESS we can reasonably choose a burn-in visually, say 5000 iterations.

```{r long-run-burn, collapse = TRUE, fig.width=8}
traceBurn <- burnAndThin(trace, burn = 5000)

# removing the burning increases the ESS
effectiveSize(traceBurn)

# autocorrelation
acfplot(traceBurn, lag.max = 60)
```

Again, given the very high ESS, we can be quite generous in our choice of the thinning.

```{r long-run-thin}
# Thinning: let's keep 1 iteration every 40
traceBurnThin <- burnAndThin(traceBurn, thin = 40)
xyplot(traceBurnThin)
```
However, let's compare the thinned and unthinnned distributions. 

```{r long-run-compare-thin, fig.width=10}
# Note that plotPosteriorDensity can take a list of mcmc.list
# It will plot the different mcmc.list by combining their elements
# Let's plot the combined unthinned trace vs the combined thinned trace.
plotPosteriorDensity(list(unthinned = traceBurn, thinned = traceBurnThin))
```

In contrast to the previous short-run, they are almost no difference between the thinned and unthinned chains. __Indeed, with such a long chain, the clumpy autocorrelation has been averaged out!__

In fact, there are several references that show that the longer (unthinned) chain usually yields better estimates of the true posterior than the shorter thinned chain, even for percentiles in the tail of the distribution. __That said, thinning can be useful for other reasons, such as memory or time constraints in post-chain processing.__

Now, we can compare whether the two independent chains, started at `theta1` and `theta2`, have converged to the same posterior distribution

```{r long-run-compare-chains}
densityplot(traceBurnThin)
```
Since the chains have converged to the same posterior, we can use the combined estimates

```{r long-run-summary, collapse = TRUE}
# the function summary combines the chains of a mcmc.list
summary(traceBurnThin)
```

Running several independent chains starting from different parts of the parameter space allows us to check whether the posterior distribution is multi-modal. If so, then we must be careful when combining the chains. For instance, an estimate of the mean computed with `summary` won't be meaningful for a parameter with a multi-modal posterior.

__By contrast, for a unimodal posteriors, combining chains is an efficient way to increase the ESS and the precision of the posterior estimates.__ Furthermore, running several "shorter" chains in parallel is faster than running one "long" chain.

Finally, let's assess the fit of the deterministic SEITL model.

```{r long-run-fit, collapse = TRUE, results = "hide", fig.height = 5, fig.width = 6}
# load data
data(FluTdC1971)

# the same initState as for the fit
initState <- c(S = 279, E = 0, I = 2, T = 3, L = 0, Inc = 0)

# by default plotPosteriorFit summarize the fit of 100 thetas sampled from the posterior
plotPosteriorFit(trace = trace, fitmodel = seitlDeter, initState = initState, data = FluTdC1971)

# alternatively,  one can plot the fit of the mean of the posterior (in this case the observation is replicated 100 times)
plotPosteriorFit(trace = trace, fitmodel = seitlDeter, initState = initState, data = FluTdC1971, posteriorSummary = "mean")

# or using the maximum a posteriori (MAP) estimate
plotPosteriorFit(trace = trace, fitmodel = seitlDeter, initState = initState, data = FluTdC1971, posteriorSummary = "max")
```

Note that the 95% credible intervals (CI) for the posterior fit under the MAP captures the highest data point. By contrast, the fit of the second peak seems quite poor, even for the MAP.

You can now go back to the [practical](mcmc_and_model_comparison.html#correlations) and look at the posterior correlations between the parameters.

# Correlations

The correlation of the posterior distribution can be investigated using `levelplot`.

```{r correlation-levelplot, fig.height = 6, fig.width = 6}
library("lattice") ## for the 'levelplot command
# levelplot doesn't accept `mcmc.list`, we pass the first `mcmc` only.
levelplot(traceBurnThin[[1]], col.regions = heat.colors(100))
```

Note the strong positive correlations (~0.8) between $R_0$ and $D_{lat}$ and between $R_0$ and $D_{inf}$. In order to explain the wide 95% CIs of $R_0$ and $D_{inf}$, let's have a look at the contact rate $\beta = R_0/D_{inf}$.

```{r correlation-beta, collapse = TRUE, warning = FALSE}
with(as.data.frame(traceBurnThin[[1]]), quantile(R0 / D_inf, probs = c(0.025, 0.25, 0.5, 0.75, 0.975)))
```
The posterior value of $\beta$ varies somewhat less than the posterior values of $R_0$ and $D_\mathrm{inf}$. Indeed, this parameter is constrained by the shape of the initial phase of the outbreak. Conversely, there are an infinite number of combinations of $R_0$ and $D_{inf}$ that lead to the same $\beta$, hence their wide 95% CIs.

A second effect that could explain the wide posterior density in $R_0$ is the very high attack rate. Indeed, once $R_0>5$ it doesn't make much difference whether $R_0$ is equal to, say, 10 or 20.

We can also note that the posterior estimate of $D_{inf} = 11$ days (95% CI: $[6-15]$) is biologically unrealistic based on previous empirical estimates. However, our approach did not include any prior information as the default `seitlDeter` fitmodel comes with uniform priors for all parameters.

In order to include previous empirical information on $D_{lat}$ and $D_{inf}$, let's modify the `dPrior` function of `seitlDeter` as follows:

```{r correlation-info-prior, eval = FALSE}
seitlDeter$dPrior <- function(theta, log = FALSE) {
  # package with truncated normal distribution
  library(truncnorm)

  logPriorR0 <- dunif(theta[["R0"]], min = 1, max = 50, log = TRUE)
  # normal distribution with mean = 2 and sd = 1 and truncated at 0
  logPriorLatentPeriod <- log(dtruncnorm(theta[["D_lat"]], a = 0, b = Inf, mean = 2, sd = 1))
  # normal distribution with mean = 2 and sd = 1 and truncated at 0
  logPriorInfectiousPeriod <- log(dtruncnorm(theta[["D_inf"]], a = 0, b = Inf, mean = 2, sd = 1))
  logPriorTemporaryImmunePeriod <- dunif(theta[["D_imm"]], min = 0, max = 50, log = TRUE)
  logPriorProbabilityLongTermImmunity <- dunif(theta[["alpha"]], min = 0, max = 1, log = TRUE)
  logPriorReportingRate <- dunif(theta[["rho"]], min = 0, max = 1, log = TRUE)

  logSum <- logPriorR0 + logPriorLatentPeriod + logPriorInfectiousPeriod + logPriorTemporaryImmunePeriod + logPriorProbabilityLongTermImmunity + logPriorReportingRate

  return(ifelse(log, logSum, exp(logSum)))
}
```

Note the choice of a truncated normal distribution since $D_{lat}$ and $D_{inf}$ must be positive.

You can now go back to the [practical](mcmc_and_model_comparison.html#informative-priors) and run a MCMC with this informative prior.

# Informative priors

Here we combine both chains with informative priors and compare the posterior distribution with the one above.

```{r info-prior-analysis, results = "hyde"}
library("coda")
# create mcmc object
traceInfo1 <- mcmc(mcmc_SEITL_infoPrior_theta1$trace)
traceInfo2 <- mcmc(mcmc_SEITL_infoPrior_theta2$trace)

# combine in a mcmc.list
trace.info <- mcmc.list(traceInfo1, traceInfo2)

# burn and thin as the chain with uniform prior (see above sections)
traceInfoBurnThin <- burnAndThin(traceInfo, burn = 5000, thin = 40)

# check that both chains converged to the same posterior
plotPosteriorDensity(traceInfoBurnThin)

# compare the effect of informative priors on the posterior distribution
plotPosteriorDensity(list(unif = traceBurnThin, info = traceInfoBurnThin))
```

$R_0$ and $D_{inf}$ have very different posterior. This is expected since there is an informative prior on $D_{inf}$ and $R_0$ is strongly correlated to $D_{inf}$. Note also that the mode of all other parameters have changed, though less than $D_{inf}$ and $R_0$. This illustrate the influence that one prior can have on the full posterior distribution.

You can now go back to the [practical](mcmc_and_model_comparison.html#model-selection).

# Model selection

```{r dic, collapse = TRUE}
# combine the two chains in a data frame
library("plyr") # needed for the 'ldply' function used in the next line
traceCombined <- ldply(traceInfoBurnThin)

# take the mean of theta
thetaBar <- colMeans(traceCombined[seitlDeter$thetaNames])
print(thetaBar)

# compute its log-likelihood
initState <- c(S = 279, E = 0, I = 2, T = 3, L = 0, Inc = 0)
logLikeThetaBar <- dTrajObs(seitlDeter, thetaBar, initState, data = FluTdC1971, log = TRUE)
print(logLikeThetaBar)

# and its deviance
dThetaBar <- -2 * logLikeThetaBar
print(dThetaBar)

# the effective number of parameters
pD <- var(-2 * traceCombined$log.likelihood) / 2
print(pD)

# and finally the DIC
dic <- dThetaBar + 2 * pD
print(dic)
```
Follow this [link](mcmc_and_model_comparison.html#posterior-predictive-checks) to go back to the practical.




