% Model fitting and inference for infectious disease dynamics

<!-- ## Lunch options -->

<!-- [Local map](http://goo.gl/XEPxch) -->

## Required software
* [R](http://cran.r-project.org) is used as the main programming language. Please install at least version: R-3.2.0.
* [RStudio](http://www.rstudio.com/products/rstudio/download/) (optional) is a popular graphic user interface (GUI).

If you are not familiar with `R`, check our [list of useful commands](slides/rintro.pdf).

## Installation of the fitR package

The simplest way to install the `fitR` is to use the `devtools` package:

```{r install, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("sbfnk/fitR@fitcourse_2015")
```

Then you can check that the installation completed successfully by loading the package into your `R` session:

```{r load, eval=FALSE}
library(fitR)
```

## Practial session

1. [Introduction](introduction.html)
2. [MCMC](mcmc.html)
3. [MCMC diagnostics](mcmc_diagnostics.html)
4. [Modelling interlude](play_with_seitl.html)
5. [MCMC and model comparison](mcmc_and_model_comparison.html)
6. [Particle MCMC](pmcmc.html)
7. [Fitting models to data using *pomp*](pomp.html)

## Extra practical session

8. [Approximate Bayesian Compuation](ABC.html)

## Examples of model fitting

[So what did happen on Tristan da Cunha?](slides/anton_talk.pdf)

[Modelling the Ebola outbreak in West Africa](slides/seb_talk.pdf)

# Navigate
Next: [Introduction](introduction.html)

</div>
