% The SEITL model in *LibBi*

```{r setup, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE, fig.path='figure/libbi/', cache.path='cache/libbi/' , fig.cap='', fig.align="center", tidy=FALSE)
```

```{r fitR, include=FALSE, cache=FALSE}
library(fitR)
set.seed(1234)
```

[Day 4 revision](revision_day_4.pdf)

# Introduction

So far, we have used self-written R code and the `fitR` package to conduct inference. This is highly inefficient. Generally, code written in machine-compiled languages such as C or C++ can be several orders of magnitude faster. Moreover, smart parallelisation of the various steps involved, as well as other techniques that accelerate computation can have a big impact on the time needed to obtain satisfactory results.

There are numerous tools around that address this, and it is difficult to come up with a comprehensive list. A popular package is [pomp](https://kingaa.github.io/pomp), which includes multiple methods and many examples from infectious disease modelling. Here, we will look at a library called *LibBi* (Library for Bayesian inference). It can be used from R via the `rbi` package.

# Installing LibBi and RBi

By far the easiest way for installing [LibBi](http://libbi.org) is using [Homebrew](https://brew.sh) (on OS X) or [Linuxbrew](http://linuxbrew.sh) (on linux). Once one of these is installed, issue the following command (using a command shell, i.e. Terminal or similar):

```{sh eval=FALSE}
brew install libbi
``` 

You'll also need the `rbi` and `rbi.helpers` packages:

```{r eval=FALSE}
install.packages('rbi')
install.packages('rbi.helpers')
```

```{r}
library('rbi')
library('rbi.helpers')
```

If you are running this on MacOS, it is probably a good idea to also run 

```{r}
options(libbi_args=list(openmp=FALSE))
```

as OpenMP is not supported by the default C++ compiler that comes with MacOS.

For an introduction to the functionalities of the `rbi` and `rbi.helpers` packages, you can have a look at the [RBi vignette](https://cran.r-project.org/web/packages/rbi/vignettes/introduction.html) and [RBi.helpers vignette](rbi_helpers.html).

# Deterministic SEITL model

Models are given in LibBi using text files that follow a specific structure. You can download our LibBi implementation of the deterministic SEITL model using

```{r}
SEITL_det <- bi_model(lines=readLines("https://sbfnk.github.io/mfiidd/bi/SEITL_det.bi"))
```

Here, `readLines` simple reads in the text file from the web site, and `bi_model` prepares it to be used with the `rbi` package.

Look at the contents of the model file by typing

```{r eval=FALSE}
SEITL_det
```

This file has the following sections:

```c
model SEITL_deter {
```

```c
  const k_erlang = 1
  const N = 1000
```

This defines two constants: the number of Erlang compartments (here: 1) and the population size (here: 1000). We can change these from within R, for example if we want more than one Erlang compartment.

```c
  dim k(k_erlang)
```

This defines a _dimension_ (i.e., the number of elements in a vector), that we’ll later use for the temporary immune compartments _T_.

```c
  state S, E, I, T[k], L, Inc
```

This defines the model states, that is variables that change over time. Note the `[k]` after `T`, specifying that `T` has dimension `k` (that is, `k_erlang` compartments).

```c
  param R0, D_lat, D_inf, alpha, D_imm, rho
```

This defines the parameter, that is variables that don’t change over time.

```c
  obs Cases
```

This defines an _observation_ variable, that is one that will be expected in the data passed to LibBi.

```c
  sub initial { 
    E <- 0
    I <- 2
    T[k] <- (k == 0 ? 3 : 0)
    L <- 0
    Inc <- 0
    S <- N - I - T[0]
  }
```

This specifies the initial states of the system. Here they are fixed, but they could be drawn from distributions, too. We start with 2 infectious individuals and 3 in the first `T` compartment.

```c
  sub parameter { 
    R0 ~ uniform(1, 50)
    D_lat ~ uniform(0, 10)
    D_inf ~ uniform(0, 15)
    D_imm ~ uniform(0, 50)
    alpha ~ uniform(0, 1)
    rho ~ uniform(0, 1)
  }
```

This specifies the prior distributions. Here, they are all uniform, just in the `fitmodel` implementation of the SEITL model.


```c
  sub transition {
```

This starts the _transition_ block, that is the process model.

```c
    inline beta = R0/D_inf
    inline epsilon = 1/D_lat
    inline nu = 1/D_inf
    inline tau = 1/D_imm
```

This defines a few auxiliary variables to be used later.

```c
    Inc <- 0
```

This resets incidence (`Inc`) at every time step.

```c
    ode { 
      dS/dt = -beta * S * I/N + (1-alpha) * tau * k_erlang * T[k_erlang - 1]
      dE/dt = beta * S * I/N - epsilon * E
      dI/dt = epsilon * E - nu * I
      dT[k]/dt =
          + (k == 0 ? nu * I : 0)
          - k_erlang * tau * T[k]
          + (k > 0 ? k_erlang * tau * T[k-1] : 0)
      dL/dt = alpha * k_erlang * tau * T[k_erlang - 1]
      dInc/dt = epsilon * E
    }
```

This defines the system of ODEs. Note the notation for `dT[k]/dt`, where the ODE depends on whether `k` is zero or greater than zero.
    
```c
  sub observation {
    Cases ~ poisson(rho * Inc)
  }
```

This defines the observation model. We use a observations distributed according to a Poisson distribution.

<!-- ```c -->
<!--   sub proposal_parameter { -->
<!--     R0 ~ truncated_gaussian(mean = R0, std = 3, lower = 1, upper = 50) -->
<!--     D_lat ~ truncated_gaussian(mean = D_lat, std = 0.2, lower = 0, upper = 10) -->
<!--     D_inf ~ truncated_gaussian(mean = D_inf, std = 0.9, lower = 0, upper = 15) -->
<!--     alpha ~ truncated_gaussian(mean = alpha, std = 0.02, lower = 0, upper = 1) -->
<!--     D_imm ~ truncated_gaussian(mean = D_imm, std = 1, lower = 0, upper = 50) -->
<!--     rho ~ truncated_gaussian(mean = rho, std = 0.006, lower = 0, upper = 1) -->
<!--   } -->

<!-- ``` -->

<!-- This defines the proposal distributions. Here, we use independent truncated normal proposals, bounded according to the uniform prior distributions. -->

# Stochastic SEITL model

Let’s load the stochastic model using
    
```{r}
SEITL_sto <- bi_model(lines=readLines("https://sbfnk.github.io/mfiidd/bi/SEITL_sto.bi"))
```

Again, you can look at the contents of the model file by typing

```{r eval=FALSE}
SEITL_sto
```

This is very similar to the deterministic model, but the `transition` block (the process model) is different:

```c
    infection ~ binomial(S, 1 - exp(-beta * I/N * timestep))
    incubation ~ binomial(E, 1 - exp(-epsilon * timestep))
    loss_infectiousness ~ binomial(I, 1 - exp(-nu * timestep))
    immunity[k] ~ binomial(T[k], 1 - exp(-k_erlang * tau * timestep))
    loss_immunity ~ binomial(immunity[k_erlang - 1], 1 - alpha)

    S <- S - infection + loss_immunity
    E <- E + infection - incubation
    I <- I + incubation - loss_infectiousness
    T[k] <- T[k] + (k == 0 ? loss_infectiousness : 0) + (k > 0 ? immunity[k - 1] : 0) - immunity[k]
    L <- L + immunity[k_erlang - 1] - loss_immunity
    Inc <- Inc + infection
```

Now, a number of random variables (defined earlier in the file as so-called `noise` variables) are first drawn from a binomial distribution and then used to update the model states.

# pMCMC with the stochastic SEITL model

We want to determine the optimal number of T compartments in the stochastic SEITL model. To do this, we run pMCMC with 400 particles and 3000 iterations (i.e., same as in the [pMCMC pratical](pmcmc.html)). But first, we prepare the Tristan da Cunha data set to be used as observation data set with `rbi`:

```{r}
library('dplyr')
tdc <- list(Cases=FluTdC1971 %>% dplyr::select(time, value=obs))
```

The `rbi` object expects data to be given as a named list, where each element corresponds to a variable of the same name, with values given by a data frame with columns `time` and `value`. Here, the observations are called `Cases`, the same as in the line specifying the observation density in the LibBi file above.

Now, we can use this data set and the model to run pMCMC. First, we generate 3000 samples where we use the prior distribution as proposal distribution, to find a good starting point for the MCMC chain. We then use this to adapt the widths of the (independent truncated normal) proposals in a series of trial runs, until they yield acceptance rates between 0.1 and 0.3.

```{r eval=FALSE}
  prep <- sample(SEITL_sto, nsamples=3000, proposal="prior", obs=tdc, 
                 end_time=max(tdc$Cases$time), nparticles=400)
  prep <- adapt_proposal(prep, min=0.1, max=0.3)
```

We can inspect the adapted model by typing

```{r eval=FALSE}
  prep$model
```

This now contains a `proposal_parameter` block at the bottom, which contains the proposal distributions.

Next, we run 3000 pMCMC iterations for 1 to 10 Erlang compartments for T.

```{r eval=FALSE}
bi <- list()
max_erlang <- 10

for (i in 1:max_erlang) {
  cat("Fitting with", i, "T compartment(s).\n")
  bi[[i]] <- sample(prep, model=fix(prep$model, k_erlang=i),
                    sample_obs=TRUE, nsamples=3000)
}
```

```{r echo=FALSE}
bi <- list()
max_erlang <- 10

for (i in 1:max_erlang) {
  cat("Fitting with", i, "T compartment(s).\n")
  bi[[i]] <- read_libbi(paste0("../rds/seitl_", i, ".rds"))
}
```

Lastly, we compute the DIC for each of these pMCMC runs.

```{r}
dic <- data.frame(compartments=integer(0), DIC=numeric(0))
for (i in 1:max_erlang) {
  dic <- rbind(dic, data.frame(compartments=i, DIC=DIC(bi[[i]])))
}
```

Let's plot the DIC as a function of the number of Erlang compartments.

```{r}
ggplot(dic, aes(x=compartments, y=DIC)) +
  geom_point() +
  geom_line()
```

Which model fits best? You can visually inspect the fit of, for example, the SEIT2L model by plotting filtered trajectories as generated by the pMCMC algorithm

```{r}
os <- summary(bi[[2]], type="obs")

ggplot(os, aes(x=time))+
    geom_line(aes(y=Median)) +
    geom_ribbon(aes(ymin=`1st Qu.`, ymax=`3rd Qu.`), alpha=0.5) +
    geom_point(aes(y=obs), FluTdC1971, color="darkred") +
    ylab("cases")
    
```

# Going further

* More information on all the methods and options and available can be found in the [LibBi manual](http://libbi.org/docs/LibBi-Manual.pdf) and the accompanying [paper](https://arxiv.org/abs/1306.3277).
* More information on the R packages can be found in the [RBi vignette](https://cran.r-project.org/web/packages/rbi/vignettes/introduction.html) and [RBi.helpers vignette](rbi_helpers.html).
* LibBi is being superseded by a new library called [Birch](https://birch-lang.org/). This allows to specify more general models than is possible in LibBi.
