% Practical session: MCMC diagnostics

```{r setup, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE, fig.path='figure/mcmc_diagnostics/', cache.path='cache/mcmc_diagnostics/' , fig.cap='', fig.align="center", tidy=FALSE)
```

```{r load, include=FALSE,cache=FALSE}
library(fitR)
data(SIR)
data(SIR_reporting)
```

[Lecture slides](slides/mcmc_slides2.pdf)

# Objectives

The aim of this session is to learn how to interpret the results of MCMC sampling, as well as to improve the performance of the sampler. More specifically, you will

1. use MCMC to sample more than one parameter
2. learn how to use `coda` for assessing and interpreting MCMC runs
3. explore strategies for improving the performance of a Metropolis-Hastings sampler

# Sampling more than one parameter

<!-- If we are fitting more than one parameter, we need to use a multivariate proposal distribution such as the multivariate Gaussian. We have included a function called `mcmcMH`in the `fitR` package which extends the Metropolis-Hastings sampler of the last session to a scenario of multiple parameters by replacing the univariate Gaussian proposal distribution of `my_mcmcMH` with a multivariate Gaussian, using the `rtmvtnorm` from the `tmvtnorm` package. In fact, the `rtmvtnorm` function calculates the probability density of a vector under a *truncated* multivariate Gaussian, that is one that can have boundaries. We will get back to this issue later. -->

<!-- The `mcmcMH` takes the same parameters as the `my_mcmcMH` function you used in the previous practical session, except that it takes a parameter `gaussian.proposal` instead of `proposal.sd`. -->

In the previous practical session, we worked through some simple examples where we sampled from a distribution that depends on just one parameter. When fitting dynamic models, however, we are often faced with the need to fit models with a multitude of parameters so that our MCMC needs to sample from a multivariate posterior distribution.

As a running example, we are going to analyse the data set `epi3`, which was generated from an SIR model with unknown basic reproduction number $R_0$ and infectious period $D_\mathrm{inf}$. 

You can visualise the `epi3` data set with

```{r epi3_traj}
plotTraj(data = epi3)
```

The first step is to write a wrapper for the log-posterior similar to `my_dLogPosterior_R0_epi1` but for more than one parameter and using the data set `epi3`  -- if you have trouble with that, you can use our [solution](epi3_wrapper.html). 

__Once you have your wrapper, take 15 minutes to modify your MCMC sampler coded in the previous session to generate samples from a multivariate posterior distribution corresponding to any `fitmodel` object. Use the new wrapper to check that your code works without bugs.__ 

Note that you should take advantage of the fact that most **R** functions, like `rnorm`, are vectorized over their arguments:

```{r vectorization_1}

set.seed(12345)
rnorm(2, mean = c(1,2), sd = c(1, 0.5))

```
is equivalent to 

```{r vectorization_2}

set.seed(12345)
c(rnorm(1, mean = 1, sd = 1), rnorm(1, mean = 2, sd = 0.5))

```

_Note that `set.seed` allows us to initialise the random number generator with the same seed in both examples. This is a useful trick to know when you want to make fully reproducible stochastic examples._

If you have difficulties to generalise your MCMC sampler to a multivariate distribution or if you didn't get to write your own MCMC sampler, take a look or just use the one from our [solution](mcmc_example_solution.html).

__Take 15 minutes to sample from the joint posterior of $R_0$ and $D_\mathrm{inf}$. In particular, try a few values for the vector of standard deviations `proposal.sd` and see how this affects the acceptance rate.__

You should notice that getting the acceptance rate to a desired value is harder when dealing with multiple parameters. Actually, there are a few strategies available for doing so, which we will learn about in this practical session.

For instance, if you experience a very low or a very high acceptance rate, it is a good idea to first consider each parameter individually, and twick the corresponding standard deviation of the proposal until a desired acceptance rate is achieved. Once this is done, you can start exploring multiple parameters at once, and carefully adjust the proposal steps to get good acceptance rates. Later, we will look at a way to automatically adapt the proposal steps on the basis of the samples taken so far.


# Diagnostics

Whenever one runs MCMC, it is important to assess its performance. It is very difficult (arguably, impossible) to be fully sure that a particular set of  MCMC runs provides reliable analysis, but, fortunately, there are ways to spot when things go wrong. We will look at assessing three aspects of MCMC: mixing, burn-in and run length


We will use functions from the **R** package `coda` to diagnose MCMC runs. __We suggest that, once you've read this section, you spend a good half hour experimenting with the Metropolis-Hastings sampler and the `coda` package. Take your time to fully understand the behaviour of the Metropolis-Hastings sampler -- this is what we will use throughout this course.__ Don't only try to get the best fit -- it is also worth experimenting with what you need to do to break the sampler, i.e. get results inconsistent with what you found before.

<!-- That is, you should try running MCMC on the `epi3` data set from different starting points, with different values for the step size (`proposal.sd`), and assess the impact on the summary statistics and plots introduced below. Experiment with burning and thinning (which we will introduce below), and see what impact they have. -->


## Summary statistics

Diagnosing MCMC usually implies a graphical assessment of the behaviour of the sampler with respect to each fitted parameter, as well as of some summary statistics, and `coda` provides a rich suite of functions to this end. We suggest that you use the commands below to assess MCMC runs with different values of the initial parameters `init.theta` and step size `proposal.sd`. If you didn't get an MCMC run to work, you can look at the output of an MCMC run that we did for you and saved in the `mcmc.epi3` object, which you can load with `data(mcmc)`:


```{r load_trace}
data(mcmc)
trace <- mcmc.epi3$trace
head(trace)
```

This run was created using the commands shown [here](mcmc_commands.html). In the following, we assume that you assigned the result of `my_mcmcMH` to a variable called `trace`, but of course you can call this anything you like.

To use `coda`, we must first convert our trace to a format that `coda` understands. To this end, we use the `mcmc` function from that package:

```{r coda_trace}
library('coda')
mcmc.trace <- mcmc(trace)
```

We can get summary statics using `summary`

```{r summary_trace}
summary(mcmc.trace)
```

This provides the following information:

1. Empirical (sample) mean
2. Empirical (sample) standard deviation
3. "Naive" standard error, that is the [standard error of the mean](http://en.wikipedia.org/wiki/Standard_error#Standard_error_of_the_mean) (adjusting for sample size).
4. Time-series standard error, which corrects the "naive" standard error for autocorrelations
5. Quantiles for each variable

We can also compute the acceptance rate (that is 1 minus the rejection rate).

```{r rejection}
acceptanceRate <- 1 - rejectionRate(mcmc.trace)
acceptanceRate
```

Lastly, we can compute the effective sample size, that is an estimate for the number of *independent* samples (taking into account autocorrelations) generated by the MCMC run.

```{r ESS}
effectiveSize(mcmc.trace)
```

## Mixing

We need to make sure that the MCMC sampler explores the parameter space *efficiently*, that is that it doesn't reject or accept too many proposals. If too many proposals are rejected, we need many simulations to generate a sufficient number of parameter samples. If too many proposals are accepted, we don't gain much information about the underlying distribution.

### Trace and density plots

Trace plots provide an important tool for assessing mixing of a chain. Density plots are smoothed histograms of the samples, that is they show the function that we are trying to explore. We can get trace and density plots for all variables in an MCMC trace using `plot`.

```{r trace, fig.width = 10, fig.height = 10}
plot(mcmc.trace)
```

_NB: the "bandwidth" indicated is automatically selected by the function to smooth the density plots._

In the trace plots, we want to try to avoid flat bits (where the chain stays in the same state for too long) or too many consecutive steps in one direction. In this case, it looks like there was a burn-in of about 1000 iterations, after which the MCMC sampler seems to mix well.

We also see that the sampler never moves beyond 2.2 for $R_0$, and never beyond 3.5 for $D_\mathrm{inf}$. To assess the reliability of our output, we should start chains with higher initial values of $R_0$ and $D_\mathrm{inf}$ and check that the sampler converges to the same estimates.

If we want to get a more detailed view of the posterior distribution (i.e., the density of the samples) around its maximum, we can cut the burn-in period out using the `burnAndThin` (this function is part of the `fitR` package but returns objects compatible with `coda` functions):

```{r trace_burned, fig.width = 10, fig.height = 10}
mcmc.trace.burned <- burnAndThin(mcmc.trace, burn = 1000)
plot(mcmc.trace.burned)
```

It is sometimes said that we are aiming for the trace to look like a [hairy caterpillar](external_ref/hairy_caterpillar.jpg). Can you see the resemblance to the plots on the left?

### Autocorrelations

Another way to check for convergence is to look at the autocorrelations between the samples returned by our MCMC. The lag-$k$ autocorrelation is the correlation between every sample and the sample $k$ steps before. This autocorrelation should become smaller as $k$ increases, i.e. samples can be considered as independent. If, on the other hand, autocorrelation remains high for higher values of $k$, this indicates a high degree of correlation between our samples and slow mixing.

```{r autocorrelations, fig.width = 10, fig.height = 5}
autocorr.plot(mcmc.trace.burned)
```

In this case, the autocorrelation drops with increasing $k$ (or "lag", the x-axis in the plot), which is a good sign. If autocorrelation persists, we can *thin* the MCMC chain, that is we discard $n$ samples for every sample that we keep. To do this, again we use `burnAndThin` and pass the number of samples $n$ to be discarded for each kept sample as `thin` argument:

```{r trace_burned_thinned, fig.width = 10, fig.height = 5}
mcmc.trace.burned.thinned <- burnAndThin(mcmc.trace.burned, thin = 5)
autocorr.plot(mcmc.trace.burned.thinned)
```

It has been argued that thinning is actually not very useful, unless one wants to reduce the amount of memory and storage space in long chains. Instead of thinning to, say, keep only 1 out every 10 samples, it is usually more efficient (in terms of the effective sample size) to just run a chain 10 times as long, but it will take 10 times more storage space.

## Burn-in

In the above example, we assessed burn-in based on a glance at the trace plots. There are several diagnostics available to make this process more systematic, such as the Geweke diagnostic (`geweke.diag`), the Heidelberger-Welch diagnostic (`heidel.diag`), and the Raftery-Lewis diagnostic (`raftery.diag`). A discussion of these is beyond the scope and purpose of this session, but if you are interested have a look at the **R** help pages for these functions (using `?geweke.diag` etc.) and the reference [below](#going-further) for more information.

A better estimate for burn-in cut-off is via the effective sample size (ESS), that is the number of independent samples equivalent to our number of autocorrelated samples. The samples in the burn-in are not very informative, and if the burn-in period is estimated to be too short this will reduce the ESS size. On the other hand, if the burn-in period is estimated to be too long, informative samples are being thrown away, again reducing the ESS. The ESS should be maximised at the optimal estimate of the burn-in. 

We can plot ESS against burn-in:

```{r test_burn_in, fig.width = 8, fig.height = 5}
plotESSBurn(mcmc.trace)
```

A good optimal burn-in length would be when the ESS has hit its maximum for all parameters, that is at around 500 here.

## Run length

It is usually difficult to tell for how long one should run the chain. If the trace of all parameters shows the "caterpillar"-like behaviour such as shown above, it is a good indication that the MCMC is efficiently sampling from a maximum in the underlying distribution, but the sampler might be stuck in a *local* maximum, and it might need more time to leave that maximum and reach other parts of the parameter space, with potentially other maxima.

There are no general rules for how long one needs to run the chain, but generally it is a good idea to start the chain from different starting points and make sure they converge to the same density plots. In fact, it is good practice to *always* run multiple chains. Secondly, one can decide in advance on the number of samples one wants from the underlying distribution, and run the chain until the effective sample size reaches that number. In practice, however, this is a somewhat arbitrary decision, and apart from careful diagnosis and running multiple chains there is not much one can do to get a reliable estimate of the necessary number of iterations.

# Improving mixing

Getting an MCMC sampler to mix well is, arguably, the most difficult and most important task for using it efficiently. We will discuss two ways for improving mixing: dealing with parameters with limited support and adaptive MCMC.

## Dealing with parameters with limited support

Often, we have situations where parameters are constrained to remain between certain values. For example, negative values do not make sense for the basic reproduction number $R_0$ or the duration of infection $D_\mathrm{inf}$, and probabilities have to stay between 0 and 1. We call the values that a parameter can, *in principle*, take the *support* of a parameter. This is different from the prior distribution which encodes our prior knowledge of likely values of a parameter among the values it can take.

If we ignore the limited support that parameters can have, we might waste MCMC steps, for example when the random draw from the proposal distribution suggests a step to a negative value of a parameter with positive support (e.g., $R_0$). In this case, proposals are rejected not because the parameter fails to match the data, but because they are in *impossible* parts of the parameter space.

### Transforming parameters

A first way for dealing with parameters with limited support applies mostly to parameters which are limited to be positive, as is often the case for *rates*, such as the ones in the SIR model. A way to force a parameter to be positive is to *transform* them to the log-scale, that is to run MCMC on $\log{(R_0)}$ instead of $R_0$. When calculating the model, we can use the exponential function to transform the parameter back.

You can load an SIR model where the parameters are defined on the log-scale using

```{r load_SIR_exp, message = FALSE, results = "hide"}
data(SIR_exp)
```

You can, again, use a wrapper function and use MCMC to explore the parameters.

```{r source_stuff, include = FALSE, cache=FALSE}
source('epi3_wrapper.r')
source('our_posterior.r')
source('our_mcmcMH.r')
```

```{r transform, message = FALSE}
my_dLogPosterior_exp_epi3 <- function(theta) {

	return(my_dLogPosterior(fitmodel = SIR_exp,
		theta = theta,
		init.state = c(S = 999, I = 1, R = 0),
		data = epi3))

}

trace <- my_mcmcMH(target = my_dLogPosterior_exp_epi3,
	init.theta = c(R0 = log(1), D_inf = log(2)),
	proposal.sd = c(0.1, 0.01),
	n.iterations = 1000)
```

__Take 10 minutes to check whether this makes a difference to acceptance rates?__

### Truncated proposal distributions

Another way for dealing with parameters with limited support is to use *truncated* proposal distributions. There are **R** packages for truncated normal and multivariate normal distributions -- these never draw proposals outside of specified limits. However, the use of such truncated distributions in the proposal kernel must be accounted for when computing the acceptance ratio. Otherwise your chain might not converge to the correct posterior distribution (look at this excellent [post](http://darrenjw.wordpress.com/tag/truncate/) by Darren Wilkinson for a detailed example).

In the `fitR` package, we have implemented a function called `mcmcMH` that uses the truncated multivariate normal random generator from the `tmvtnorm` package to draw proposals. Of course the function accounts for this asymmetric proposal to calculate the acceptance ratio. The limits of the support of the proposal distribution can be specified with the `limits` argument to `mcmcMH`, in which you can specified `lower` and/or `upper` limits. For example, to get a lower bound of 0 for $R_0$ and $D_\mathrm{inf}$, we could write

```{r posterior_epi3}
my_dLogPosterior_epi3 <- function(theta) {

	return(my_dLogPosterior(fitmodel = SIR,
		theta = theta,
		init.state = c(S = 999, I = 1, R = 0),
		data = epi3))

}
```

```{r truncate, message = FALSE}
trace <- mcmcMH(target = my_dLogPosterior_epi3,
	init.theta = c(R0 = 1, D_inf = 2),
	proposal.sd = c(0.1, 0.01),
	n.iterations = 1000,
	limits = list(lower = c(R0 = 0, D_inf = 0)))
```

__Take 10 minutes to try a proposal distribution with a wide proposal distribution, and with or without limits. Do the limits make a difference to the acceptance rate?__

## Adapting the proposal distribution

The best proposal distribution is the one that best matches the target distribution. While we cannot know this in advance, we can use trial MCMC runs to learn about the target distribution, and use this information to come up with a better proposal distribution. This does mean, however, that we waste computational time in the discarded trial runs, and this method needs to be applied carefully.

In addition to truncated proposal distributions, our function `mcmcMH` also implements some adaptive features. In particular, it calculates the *empirical* covariance matrix, that is the covariance matrix as it appears from the distribution of accepted proposals. This is returned as `covmat.empirical`:


<!-- ```{r adapt, eval=FALSE, message = FALSE, warn = FALSE} -->
<!-- trace <- mcmcMH(target = my_dLogPosterior_epi3, -->
<!-- 	init.theta = c(R0 = 1, D_inf = 2), -->
<!-- 	proposal.sd = c(0.1, 0.01), -->
<!-- 	n.iterations = 1000, -->
<!-- 	limits = list(lower = c(R0 = 0, D_inf = 0))) -->
<!-- ``` -->

```{r print_covmat}
trace$covmat.empirical
```

The diagonal elements of this matrix tell you about the typical variance of accepted steps and thus gives you an indication of an appropriate step size. The off-diagonal elements tell you about *correlations* in the parameters. In this case, there is positive correlation between $R_0$ and $D_\mathrm{inf}$, that is greater values of $R_0$ in accepted proposals tend to coincide with greater values of $D_\mathrm{inf}$. __Can you think of why that is the case?__ We will come back to the issue of correlations and their impact on assessing model fits tomorrow.

Instead of manually adjusting the proposal distribution as we have done so far, we can automate this process. If this is done carefully, the samples generated by the MCMC converge to the correct distribution. We have implemented this in the `mcmcMH` function. This takes three parameters relating to adaptive MCMC: `adaptive.size.start`, `adaptive.shape.start` and `adapt.size.cooling`.

The idea is the following: we let the MCMC run for a while and monitor the acceptance rate. After `adapt.size.start` iterations, we start adapting the *size* of the proposal distribution, that is we scale it to smaller/larger  steps depending if the acceptance is too small/large. This is done until `adapt.size.shape` proposals have been accepted, at which point we take the empirical covariance matrix and start adapting the *shape* of the proposal distribution to it. Over time, we must make fewer and fewer changes to the size, because if we kept adapting the size we would break the Metropolis-Hastings algorithm. This is regulated by `adapt.size.cooling`. The closer this is to 1, the slower we stop adapting the size (and, accordingly, the longer we have to run the chain).

Try a run of adaptive MCMC using

```{r adapt2, message = FALSE, warning = FALSE}
trace <- mcmcMH(target = my_dLogPosterior_epi3,
	init.theta = c(R0 = 1, D_inf = 2),
	proposal.sd = c(1, 0.5),
	n.iterations = 5000,
	adapt.size.start = 100,
	adapt.shape.start = 500,
	adapt.size.cooling=0.999,
	limits = list(lower = c(R0 = 0, D_inf = 0)))
```

Don't worry if you see warnings -- this could be due to numerical errors in the ODE solver of `deSolve`, which can lead to negative data points, which in turn produces `NaN`s ("Not a Number) when evaluating the likelihood.

__Take note of the acceptance rate when the adaptive sampler starts adapting the size first, and then the shape. Have a look at the trace plots (see above) and the effect that adaptation has. Does it work in ensuring good mixing?__

# Going further

To read more about MCMC diagnostics, have a look at Cowles & Carlin's excellent [review](external_ref/Cowles96_MCMCconvergence.pdf).

You can also try to fit a model with reporting to the `epi4` data set. This is a model with 3 parameters (the basic reproduction number $R_0$, the infectious period $D_\mathrm{inf}$ and the reporting rate $RR$) and it can loaded with

```{r load_SIR_rep}
data(SIR_reporting)
```

Have a look at the point observation probability density function

```{r look_loglike, eval = FALSE}
SIR_reporting$dPointObs
```

Can you fit this to the `epi4` data set?
