% Solution: the particle MCMC

```{r setup, echo=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.path = "figure/pmcmc_solution/", cache.path = "cache/pmcmc_solution/", fig.cap = "", fig.align = "center", message = FALSE, tidy = TRUE, warning = FALSE)
```

```{r fitR, echo=FALSE, cache=FALSE, results="hide"}
library("fitR")
source(here::here("scripts", "snippets", "our-smc.r"))
set.seed(1234)
```

# Run a particle filter

The particle filter returns a Monte-Carlo estimate of the log-likelihood and, as every Monte-carlo estimate, its precision depends on the number of particles.

If you have too few particles then you will have a highly variable estimate of the log-likelihood and this will make the exploration of the likelihood surface quite imprecise. In addition, you might experience particle depletion (if you don't know what that means just try to run a particle filter with a single particle).

If you have too many particles, then you will have an accurate estimate of your log-likelihood but it will be very time consuming so inefficient in practice.

In addition, the variability of the estimate might also depend on the region of the parameter space. For instance, in some region you might experience significant variability with 100 particles whereas in another region it might be fairly stable.

You can now return to the [practical](pmcmc.html#calibrate-the-number-of-particles) and try to think on a way to calibrate the number of particles.

# Calibrate the number of particles

Ideally we want enough particles to obtain a fairly stable estimate of the log-likelihood in a reasonable computational time.

A simple calibration approach consists in plotting the mean, standard deviation and computational time of the log-likelihood estimate as a function of the number of particles. Although several parameters can be tested, we will use a `theta` close to the mean posterior estimate of the deterministic fit since it is likely to be the region of the parameter space we want to explore with the pMCMC.

```{r calibration-SMC, eval = FALSE, tidy = FALSE}
knitr::read_chunk(here::here("scripts", "snippets", "calibrate-smc.r"))
```

We ran this calibration algorithm and obtained the following results:

```{r plot-calibration, echo=FALSE, fig.height=6, fig.width=8}
data(calibrateSmc)

## estimate time for 10,000 iterations from time for 100 iterations
calibrateSmc <- mutate(calibrateSmc,
  days10000 = days * 100
)
calibrateSmc <- select(calibrateSmc, -days)

dfPlot <- pivot_longer(calibrateSmc, c(-nParticles), names_to = "variable")

dfPlot <- mutate(dfPlot,
  variable = recode(variable,
    days10000 = "time 10000 iter (in days)",
    propDepleted = "prop. of samples with particle depletion"
  )
)

p <- ggplot(dfPlot, aes(x = nParticles, y = value)) +
  facet_wrap(~variable, scales = "free_y") +
  geom_line() +
  geom_vline(xintercept = 408, col = "red", linetype = "dashed") +
  theme_bw() +
  xlab("number of particles")
print(p)
```

We can see how, with increasing number of particles, the standard deviation of the log-likelihood estimate decreases whilst the run time increases. Ideally we want to balance between a stable enough estimate of the log-likelihood to ensure good mixing and computation time. Various studies have shown that this is broadly the case at values of the standard deviation between 1 and 3. TODO INSERT REF The mean and standard error of the log-likelihood estimate seem to stabilize around 400 particles (red line). By contrast, the computational time increases linearly with the number of particles. Using 400 particles looks optimal if we want a stable estimator while minimizing the computational time. Note however that it will take around 6 days to run 10000 times the particle filter with 400 particles. Since every step of the pMCMC requires to run a particle filter, that means that it will take 6 days to run 10000 iterations of the pMCMC. This is why you generally need a scientific computing cluster to run a pMCMC. 

You can now return to the [practical](pmcmc.html#run-pmcmc) and set your pMCMC with 400 particles.

# Setting the pMCMC

```{r set-pmcmc, eval=FALSE, tidy = FALSE}
knitr::read_chunk(here::here("scripts", "snippets", "set-pmcmc.r"))
```

You can now go back to the [practical](pmcmc.html#run-a-pmcmc) and try to run 	pMCMC with those settings.

# Analyse a pMCMC with 50 particles

Here is an example of analysis of our 5 chains of 3000 iterations with 50 particles.

```{r analyse-n8, collapse=TRUE, fig.height=7, fig.width=10}
# load traces
data(pmcmcSeit4lInfoPrior)

# combine into a `mcmc.list` object
library("coda")
trace <- mcmc.list(lapply(pmcmcSeit4lInfoPrior8, function(chain) {
  mcmc(chain$trace)
}))

# acceptance rate is way below the optimal 23%
1 - rejectionRate(trace)

# accordingly, the combined ESS is a bit low
effectiveSize(trace)

# Let's have a look at the traces
library("lattice")
xyplot(trace)
```

The burn-in period looks relatively short, this is because we started the chains with a "good" `initTheta`. That said, since we have relatively short runs it might be useful to carefully choose the burn-in period with `plotEssBurn`.

```{r analyse-n50-burn, collapse=TRUE, fig.height=7, fig.width=10}
# this can take some time as we have 5 chains
plotEssBurn(trace)

# Actually, it looks like no burn-in is needed
# What about autocorrelation
acfplot(x = trace, lag.max = 50)

# There is substantial autocorrelation but we can't thin too much since the chains are quite short.
# So let's keep 1 iteration every 20
traceThinN50 <- burnAndThin(trace, thin = 20)

# Finally we can plot the posterior density
densityplot(x = traceThinN50)
```

The 5 chains converged to somewhat different posterior distributions, with different shapes and modes. This is particularly evident for the posterior of $R_0$, which looks bimodal. In addition, the ESS is too small to have smooth posterior distributions.

With so few particles, the likelihood estimate is very noisy and the exploration of the parameter space is not efficient. This is why we have a lot of autocorrelation and small ESSs. That said, the theoretical properties of the pMCMC guaranty that the chain will converge to the true posterior, even with 1 particle. Of course, this will take a lot of iterations so in practice it might be more efficient to spend more time computing the likelihood (i.e. having more particles) in order to reduce the number of iterations.

You can now return to the [practical](pmcmc.html#analyse-a-pmcmc-with-400-particles) and analyse a pMCMC with much more particles that we ran for you.


# Analyse a pMCMC with 400 particles

Here is an example of analysis of our 5 chains of 3000 iterations with 400 particles.

```{r analyse-n64, collapse=TRUE, fig.height=7, fig.width=10}

# combine into a `mcmc.list` object
library("coda")
trace <- mcmc.list(lapply(pmcmcSeit4lInfoPrior64, function(chain) {
  mcmc(chain$trace)
}))

# acceptance rate is near optimal
1 - rejectionRate(trace)

# Note that the combined ESS is 2 times higher than with 50 particles
effectiveSize(trace)

# Let's have a look at the traces
library("lattice")
xyplot(trace)
```

As in the analysis with 50 particles, the burn-in period is relatively short. However, with 400 particles the chains mix much better. Here again we decide to carefully choose the burn-in period with `plotEssBurn`.

```{r analyse-n400-burn, collapse=TRUE, fig.height=6, fig.width=10}
# Actually, it looks like no burn-in is needed:
plotEssBurn(trace)

# What about autocorrelation?
acfplot(x = trace, lag.max = 50)

# Autocorrelation decreases much more quickly than with 50 particles
# Let's keep 1 iteration every 20
traceThinN400 <- burnAndThin(trace, thin = 20)

# Let's plot the posterior densities
densityplot(x = traceThinN400)
```

All 5 chains seems to have converged to the same posterior, which are smoother than with 50 particles. Let's compare the combined posterior densities with that obtained with 50 particles

```{r n400-n50, fig.height=7, fig.width=7}
plotPosteriorDensity(list(n50 = trace.thinN50, n400 = traceThinN400))
```

__Although the posterior distributions are similar, those with 400 particles are smoother and more representative thanks to higher ESS.__ Note the different location of the mode of $R_0$, which is shifted to the left with 50 particles. This is because 1 of the 5 chains with 50 particles shows a posterior with much lower $R_0$ (see figure above), whereas the remaining 4 seems to have converged to the same distribution as the pMCMC with 400 particles.

Finally, note that the log-likelihood is overestimated with 50 particles, which can be problematic for model selection as we would overestimate the fit of the model.

__Overall, this analysis confirms that the pMCMC works even with 50 particles but that it will require much more iterations to achieve the same posterior as the pMCMC with 400 particles.__ Although the latter takes more time at each iteration, it provides more better samples on short-runs. A good strategy is therefore to run many short chains in parallel with 400 particles. The chains start at different `initTheta` near the mode of the deterministic posterior, and are then combined to increase the overall ESS.

You can now return to the [practical](pmcmc.html#stochastic-vs-deterministic-fit) and proceed to the last section of this session.

# Stochastic vs deterministic fit

Here we compare the combined traces of the deterministic SEIT4L model (2 chains of $10^5$ iterations) with those obtained with the stochastic version (5 chains of $3000$ iterations). Both analysis have assumed informative priors for $D_\mathrm{lat}$ and $D_\mathrm{inf}$ .

```{r sto-deter, collapse=TRUE, fig.height=7, fig.width=7}
# load, burn and thin the deterministic fit

# create mcmc object
library("coda")
trace1 <- mcmc(mcmc_SEIT4L_infoPrior_theta1$trace)
trace2 <- mcmc(mcmc_SEIT4L_infoPrior_theta2$trace)

# combine in a mcmc.list
trace <- mcmc.list(trace1, trace2)

# burn and thin as the chain with uniform prior (see above sections)
traceDeter <- burnAndThin(trace, burn = 5000, thin = 40)

# compare posterior density
plotPosteriorDensity(list(deter = traceDeter, sto = traceThinN400))
```

Overall, the posterior distributions are quite different. This is especially true for $R_0$ and $D_{imm}$.
In addition, the discrepancy in the posterior distribution of the log-likelihood seems to indicate that the stochastic model fits much better.
We can quantify this by computing the DIC of the stochastic SEIT4L model.

```{r sto-deter-DIC, collapse=TRUE}
# combine all traces in a data frame
library("plyr")
traceCombined <- ldply(traceThinN400)

# take the mean of theta
thetaBar <- colMeans(traceCombined[seit4lStoch$thetaNames])
print(thetaBar)

# compute its log-likelihood
initState <- c("S" = 279, "E" = 0, "I" = 2, "T1" = 3, "T2" = 0, "T3" = 0, "T4" = 0, "L" = 0, "Inc" = 0)
logLikeThetaBar <- my_particleFilter(seit4lStoch, thetaBar, initState,
  data = fluTdc1971, nParticles = 400
)
print(logLikeThetaBar)

logLikeThetaBarDeter <- dTrajObs(seit4lDeter, thetaBar, initState, data = fluTdc1971, log = TRUE)
print(logLikeThetaBarDeter)

# and its deviance
dThetaBar <- -2 * logLikeThetaBar
print(dThetaBar)

# the effective number of parameters
pD <- var(-2 * traceCombined$logLikelihood) / 2
print(p.D)

# and finally the DIC
dic <- dThetaBar + 2 * pD
print(dic)
```

In the previous session, we found that the DIC of the deterministic SEIT4L model was equal to 265. The difference of 20 indicates that the stochastic model should strongly be preferred to the deterministic model.

We can visually check this result by plotting the posterior fit of each model:

```{r sto-deter-fit, results="hide", fig.height=4, fig.width=5}
# take the mean posterior estimates of the deterministic model
x <- summary(traceDeter)
thetaBarDeter <- x$statistics[seit4lDeter$thetaNames, "Mean"]

plotFit(seit4lStoch, thetaBar, initState, data = fluTdc1971, nReplicates = 1000)

plotFit(seit4lDeter, thetaBarDeter, initState, data = fluTdc1971, nReplicates = 1000)
```

Despite the fact that the deterministic model seems to better capture the first peak of the epidemic, the stochastic model better explains the variability of the observed time-series. In particular, the 95% CI of the stochastic model captures almost all the observed data points, even during the first peak.


You've already finished? Why not [go further](pmcmc.html#going-further)?
