---
title: "Modelling of past and current Ebola outbreaks"
---

```{r setup, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.path = "figure/ebola_project/", cache.path = "../cache/ebola_project/", fig.cap = "", fig.align = "center", message = FALSE, tidy = TRUE, warning = FALSE)
```

```{r fitR, cache=FALSE, echo=FALSE}
library(fitR)
set.seed(1234)
```

# Objectives

* Design and fit a simple SEIR model to one of the Ebola outbreak below.
* Estimate the change, if any, of the reproduction number ($R_t$) over time.
* Discuss your findings and the limitations of your model. 

# Data

We provide 4 data sets, the first one corresponds to the first ever reported outbreak of Ebola that occurred in Yambuku, DRC, in 1976. It is a complete epidemic of about 300 cases. 

* [Download the daily incidence time series](data/Ebola_Yambuku_1976.csv)

The 3 other data sets correspond to the ongoing Ebola outbreak in the 3 most affected countries of West Africa (Guinea, Liberia and Sierra Leone). 
The dynamics in the 3 countries is somewhat different and you will compare your estimates of $R_t$ across groups. Make sure you don't pick the same data set as another group.

The data were extracted from [WHO website](http://www.who.int/csr/disease/ebola/situation-reports/en/?m=20141126) and compiled the number of Confirmed and Probable cases up to 23 November 2014.

* [Download the weekly incidence time series](data/Ebola_West_Africa_2014_WHO.csv)

# Model

Modify one of the model provided during the practicals (SIR or SEITL) to include a time varying reproduction number. You will also need to choose an observation model, priors for the parameters as well as initial conditions for your model.

# Bibliography

You might find the following reference helpful to understand the epidemiological context of these outbreaks. The reference also contains useful empirical estimates to parametrize your model (e.g. priors, fixed parameters).

* [Breman, J. G., et al. "The epidemiology of Ebola hemorrhagic fever in Zaire, 1976." Ebola virus haemorrhagic fever (1978): 103-124.](external_ref/Breman_1978.pdf)
* [WHO Ebola Response Team. "Ebola virus disease in West Africa—the first 9 months of the epidemic and forward projections." N Engl J Med 371.16 (2014): 1481-95.](external_ref/WHO_Ebola_Response_Team_2014_New_England_Journal_of_Medicine.pdf)

