% Model fitting and inference for infectious disease dynamics

## Required software
* [R](http://cran.r-project.org) is used as the main programming language. Please install at least version: R-3.1.0.
* [RStudio](http://www.rstudio.com/products/rstudio/download/) (optional) is a popular graphic user interface (GUI).

## Installation of the fitR package

The simplest way to install the `fitR` is to use the `devtools` package:

```{r eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("sbfnk/fitR")
```

Then you can check that the installation completed successfully by loading the package into your `R` session:

```{r}
library(fitR)
```

<div>

## Contents

1. [Introduction](introduction.md)
2. [MCMC](mcmc.md)
3. [MCMC diagnostics](mcmc_diagnostics.md)
<!-- 4. [Modelling interlude](play_with_seitl.md) -->
<!-- 5. [MCMC and model comparison](mcmc_and_model_comparison.md) -->
<!-- 6. [Particle MCMC](pmcmc.md) -->
<!-- 7. [Approximate Bayesian Computation](ABC.md) -->

4. Modelling interlude
5. MCMC and model comparison
6. Particle MCMC
7. Approximate Bayesian Computation

# Navigate
Next: [Introduction](introduction.md)

</div>
