# Model fitting and inference for infectious disease dynamics

## Required software
* [R](http://cran.r-project.org) is used as the main programming language. Please install at least version: R-3.1.0. 
* [RStudio](http://www.rstudio.com/products/rstudio/download/) (optional) is a popular graphic user interface (GUI).
* [Tracer](http://tree.bio.ed.ac.uk/software/tracer) is a program for analysing the trace files generated by Bayesian MCMC runs.

## Installation of the fitcourseR package

The simplest way to install the `fitcourseR` is to use the `devtools` package:

```{r eval=FALSE}
# install.packages("devtools")
library(devtools)
install_github("sbfnk/mfiidd/fitcourseR")
```

Then you can check that everything went fine by loading the package into your `R` session:

```{r}
library(fitcourseR)
```

## Contents

1. [Overview of fitcourseR](fitcourseR.md)
2. [My first fitmodel](first_fitmodel.md)
3. [Data and likelihood](data_likelihood.md)
4. [Run a MCMC](mcmc.md)
5. [Tristan da Cunha outbreak](play_with_seitl.md)
6. [Run a SMC](smc.md)
7. [ABC inference](ABC.md)


# Navigate
Next: [Overview of fitcourseR](fitcourseR.md)